---
title: "Demand Prediction For Bike Sharing Systems"
author: "Ahmed ElSabbagh"
date: "2022-11-26"
output:
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Project Description

## Introduction

+ Bike Sharing Systems (BSS) have a very widely used method of transportation for many people worldwide.

+ They have several advantages:
    + Non-polluting.
    + Circumvents traffic congestion.
    + Excellent for last-mile connections.
    + Have a convenient payment system.
    
## Motivation

+ Bikes are parked and rented from fixed automated bike locations.
+ BSSs have a finite number of bikes and a finite number of station spaces.
+ Not all stations have an equal amount of demand at any given time of the day.
+ Bikes need to be relocated from low demand to high demand stations.
+ Companies need to have estimate the demand and available bikes at each station.


## Objectives

The objective is to create multivariate regression models capable of estimating the following

+ **Demand prediction:** Number of bikes needed at each individual station.
+ **Availability prediction:** Number of bikes available at each individual station.

#  Literature Review

# Dataset

## Lyft/Baywheels

The data was provided by Lyft, the owner of the BSS company Baywheels (formally GoBike) operating in San Francisco Bay Area.

+ The data is available as far as 2017 before Lyft's acquisition of GoBike.
+ The data used however is limited to 2021 and 2022 because:
    + The disruption caused 2020 COVID pandemic.
    + Potential business growth gap between 2019 and 2021.

## Weather Data

+ Weather data is very relevant in determining the demand.

+ The weather data for San Francisco Bay Area is provided by Meteostat through a Python API.


## Dataset Description

### Size:

+ **Original Number of Trips:** 4053524 trips .
+ **Final Number of Trips:** 3463728 trips. (subject to change)
+ **Number Of Stations**: 532 
+ **Period:** From January 2021 to September (as far as when I am finished) 2022 (2023?)

### Relevant Columns:

+ Station Name (start/end)
+ Time (start/end)
+ Coordinates (start/end)
+ Ride ID

# Cleaning Process

## Cleaning Process

+ **Station Standardization:** 

    + Stations in a particular street don't always have the same coordinates.
    + Many stations don't have a standardized name or ID, making them unidentifiable.
    + **Solution To Reduce Data Loss**: 
        + Use a single coordinate for any identifiable stations.
        + Approximate the closest standard station to any trip.
        + If the closest station is less than 500 meters away, keep the trip, otherwise, drop.

## Cleaning Process

+ **Same Station Trip:**
    + Several trips usually take place in several minutes, with the end station being the same as start stations.
    + This could be a result of users trying out the system or changed minds.
    + To prevent redundancy, any same station trip with duration less than 4 minutes will be removed.
    
## Cleaning Process
+ **Clustering**:
    + Helps reduce weather data size by:
        + Approximating areas closest to each other.
        + Approximating the weather conditions for each cluster.

## Cleaning Process

```{r, echo=FALSE, results='asis'}
htmltools::tags$iframe(title = "My embedded document", src = "Plots/clusters.html")
```

## Demand and Availability

**Definitions**

+ Demand: Number of trips starting at a particular station.
+ Availability: Number of trips ending at a particular station.
+ "Availability" does not indicate actual number of bikes available, but it indicates how many bikes finished their trips there, which when compared with demand should provide a good idea about the available bikes.
+ Therefore, demand will be the main focus, while availability will be taken as secondary.

## Demand
```{r, echo=FALSE, results='asis'}
htmltools::tags$iframe(title = "My embedded document", src = "Plots/Daily Demand Overall.html")
```

## Demand and Availability

```{r, echo=FALSE, results='asis'}
htmltools::tags$iframe(title = "My embedded document", src = "Plots/Hourly Demand vs Availability Overall (sample 5 days in September).html")
```

## Demand and Availability

```{r, echo=FALSE, results='asis'}
htmltools::tags$iframe(title = "My embedded document", src = "Plots/Hourly Demand vs Availability at Market St at 10th St.html")
```

## Demand Spread

```{r, echo=FALSE, results='asis'}
htmltools::tags$iframe(title = "My embedded document", src = "Plots/Hourly Demand Spread.html")
```

## Demand Spread

```{r, echo=FALSE, results='asis'}
htmltools::tags$iframe(title = "My embedded document", src = "Plots/Monthly Demand Spread.html")
```

## Demand Spread

```{r, echo=FALSE, results='asis'}
htmltools::tags$iframe(title = "My embedded document", src = "Plots/Demand Weekday vs Weekend.html")
```
